# Hypothesizing {#sec-hypothesizing-chapter}

Most entrepreneurs make the same mistake: they treat their guesses like facts.\
They “know” customers want their product, “know” the market is huge, and “know” the price is right — until reality proves otherwise.

The antidote is the humble **hypothesis**.

## From Guessing to Hypothesizing

A hypothesis is not just a fancy word for guess.\
It’s a **testable statement** that turns an urgent unknown into something you can check with evidence.

-   A guess says: *“People will love our product.”*\
-   A hypothesis says: *“At least 30% of gym members offered a sample will purchase one unit at \$2.50 within the next week.”*

The difference? A hypothesis can be wrong in a useful way. If the evidence doesn’t support it, you’ve learned something specific and can adjust.

## Why Hypotheses Matter

Without hypotheses, entrepreneurs fall into dangerous habits:

-   **Confirmation bias.** Seeing what you want to see in customer reactions.\
-   **Vague victories.** Believing “people liked it” means success.\
-   **Wasted effort.** Running tests that generate noise instead of insight.

With hypotheses, your learning becomes sharp. Each test produces a clear signal: supported or not. That clarity saves time, money, and frustration.

## The Anatomy of a Good Hypothesis

A strong hypothesis is:

1.  **Focused.** It addresses one urgent unknown, not everything at once.\
2.  **Measurable.** It specifies what outcome you’ll look for.\
3.  **Time-bound.** It sets a horizon for testing.\
4.  **Falsifiable.** It could be proven wrong with real evidence.\
5.  **Grounded.** It connects to something real you’ve observed, heard, or reasoned about — not just imagination.

::: {.callout-tip appearance="minimal"}
“If we \[*take action*\], then \[*this measurable result*\] will happen within \[*timeframe*\].”
:::

::: {.callout-note icon="false" title="Mini-Case: QuickBite Food Truck"}
QuickBite is a startup run by two friends who want to serve healthy grain bowls from a food truck near office parks.

They’ve identified several urgent unknowns:

-   Will busy professionals actually choose a grain bowl over faster options like pizza or burgers?\
-   Is the \$12 price point viable, or will it scare customers away?\
-   Are lunchtime crowds large enough to sustain daily sales?

Instead of treating these as vague worries, QuickBite writes a hypothesis:

**Hypothesis:**\
*“If we park the truck outside TechHub Plaza for five weekdays, then at least 40 different customers will purchase a grain bowl at \$12.”*

Why is this strong?

-   **Focused:** It’s about one unknown — willingness to buy at \$12.\
-   **Measurable:** Sales numbers are easy to track.\
-   **Time-bound:** One week.\
-   **Falsifiable:** If fewer than 40 customers buy, the hypothesis is not supported.\
-   **Grounded:** It comes from prior conversations with busy professionals who said they wanted something healthier than burgers but faster than sit-down meals.

The test gives them a clear signal: if the hypothesis fails, they may need to adjust price, location, or menu — but either way, they’ll learn something specific instead of guessing.\
The grounding makes the test worth running. Even if it fails, QuickBite will have learned something that connects back to what real people said or did.
:::

## Why Grounding Matters {#sec-groundedness-sec}

Without grounding, you’re not really testing a hypothesis. You’re just plucking guesses from an infinite list and running them through the motions. Grounding reassures you the test is worth running in the first place. If you can’t point to an observation or insight behind your statement, that’s your cue to step back and explore before you test.

Groundedness also connects back to [uncertainty](ch02-uncertainty.qmd).\
In low-uncertainty situations, you often have plenty of prior knowledge, data, or industry norms to ground a hypothesis right away.

But in high-uncertainty situations, you don’t. Hypotheses there will feel flimsy until you do some [exploration](ch10-exploring.qmd) — talking to people, noticing patterns, gathering small clues. That’s not a flaw. It’s simply a reminder that in uncertain spaces, exploration is the work that makes hypotheses possible.

```{r}
#| label: fig-groundedness-threshold
#| fig-cap: "The Groundedness Threshold: when uncertainty is high and grounding is low, you should explore before testing hypotheses."
#| fig-width: 7
#| fig-height: 5

library(ggplot2)

# Make a small grid for plotting
df <- expand.grid(
  Groundedness = seq(0, 1, length.out = 100),
  Uncertainty = seq(0, 1, length.out = 100)
)

# Define threshold line: simple diagonal (Uncertainty = 1 - Groundedness)
df$Zone <- ifelse(df$Groundedness < df$Uncertainty, "Explore", "Hypothesize")

ggplot(df, aes(x = Groundedness, y = Uncertainty, fill = Zone)) +
  geom_tile(alpha = 0.7) +
  scale_fill_manual(
    values = c("Explore" = "#E6F7FA", "Hypothesize" = "#38B2AC")
  ) +
  geom_abline(slope = 1, intercept = 0, color = "#007BA7", linewidth = 2, linetype = "solid") +
  labs(
    x = "Grounding (Low → High)",
    y = "Uncertainty (Low → High)",
    fill = "Action",
    title = "Groundedness Threshold for Hypotheses"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#1A202C"),
    axis.title = element_text(face = "bold"),
    axis.text = element_blank(),   # alternative way (also removes values)
    legend.position = "top"
  )
```

When uncertainty is low, you don’t need much direct grounding — precedent, data, and experience already supply enough information to make hypotheses useful.\
As uncertainty rises, the bar for groundedness rises too. High-uncertainty domains demand hypotheses anchored in direct observation, exploration, or customer evidence.\
The diagonal line marks the threshold: above it, you’re just guessing; below it, your hypotheses are worth testing.\
<!-- That’s why hypotheses in high-uncertainty domains must rest on evidence you’ve gathered yourself, or they risk being nothing more than guesses. -->

## Common Pitfalls

Entrepreneurs often misfire when writing hypotheses. Watch out for:

-   **Too broad.** *“Customers like healthy drinks.”* (Not testable in a week.)\
-   **Too vague.** *“People are interested.”* (What does “interested” mean?)\
-   **Too certain.** *“Everyone will prefer TechHub Plaza.”* (Absolute claims leave no room to learn.)\
-   **Unmeasurable outcomes.** *“The brand will gain awareness.”* (How will you measure “awareness”?)\
-   **Ungrounded guesses.** *“College students will definitely buy this.”* (Says who? If you can’t point to evidence or observation, you’re testing blind.)\
    A good hypothesis doesn’t predict perfection. It creates a learning moment.

## Abduction: Where Hypotheses Come From

Not every hypothesis starts with a neat theory.\
Sometimes you just notice something odd — a surprising comment, a pattern in the data, a behavior that doesn’t fit.

That’s where **abduction** comes in. Abduction is the logic of “best explanation.” It’s how you take a puzzling observation and propose a plausible story that could make sense of it.

Example: QuickBite’s team noticed that several customers asked if they sold smoothies, even though the truck specialized in grain bowls. That surprise led to an abductive leap:\
*“Maybe what these customers really want isn’t lunch, but quick portable nutrition.”*

From there, QuickBite could write a hypothesis:\
*“If we add a protein smoothie option for one week, then at least 20% of customers will choose it over the grain bowl.”*

Abduction doesn’t prove anything. It gives you a starting point — a new angle worth testing.\
And it grounds the hypothesis in something observed rather than imagined, keeping you above the threshold where testing is worth the effort.

## From Urgent Unknowns to Hypotheses

Think back to your urgent unknowns. Each one is a question waiting to be reframed.

#### Profitability

-   Urgent Unknown: *“Will customers pay enough to cover costs?”*\
-   Hypothesis: *“At least 25% of surveyed target customers will say they would purchase at \$12.”*

#### Channel

-   Urgent Unknown: *“Is retail or direct-to-consumer the better channel?”*\
-   Hypothesis: *“In a two-week trial, direct sales at gyms will produce at least 3x more unit sales than shelf placement in one convenience store.”*

## Connecting Hypotheses to Bayesian Updating

Back in @sec-belief-revision-chapter, you learned how to update beliefs the Bayesian way:

-   Start with a prior belief.\
-   Gather new evidence.\
-   Revise the probability up or down.

Hypotheses are the missing bridge. They *create* the prior.\
When you write a clear, grounded, testable hypothesis, you’re not just making a guess — you’re defining what counts as evidence and how you’ll learn from it.

That’s why every hypothesis is a launchpad for updating. Without it, you’re just collecting opinions. With it, you’re building knowledge.

## The Payoff

Writing hypotheses forces discipline. It moves you from hand-waving to clarity.\
It gives you a specific outcome to check — one that keeps you honest about whether you’re really learning.

Every urgent unknown deserves a grounded hypothesis. In @sec-exploration-chapter, we’ll explore how to *add grounding through exploration* — gathering the evidence that shrinks uncertainty and guides your next move.

## Try This: Write a Proper Hypothesis

::: {.callout-tip appearance="minimal"}
Pick one urgent unknown you identified in the last chapter.

1.  Write it down as a plain question: *“Will customers pay enough to cover costs?”*\
2.  Use the template: *“If we \[take action\], then \[this measurable result\] will happen within \[timeframe\].”*\
3.  Fill it in. Example: *“If we sell our product in one location for two weeks, then at least 50 unique customers will buy at least one unit.”*\
4.  Check it against the five tests: Focused, Measurable, Time-bound, Falsifiable, Grounded.\
5.  Now place your hypothesis on the uncertainty–groundedness chart:
    -   If uncertainty is low, little direct grounding is required
    -   If uncertainty is high, your hypothesis must rest on strong observations or exploration
6.  Revise until it both passes all five tests and sits above the threshold line in the chart.

By the end, you’ll have turned a foggy unknown into a clear, grounded hypothesis you can actually test.
:::

## Example Threshold Check

::: {.callout-note appearance="minimal"}
-   Above the threshold (high uncertainty, low grounding): “If we **run a TikTok ad**, then 10,000 people will buy in a week.”
    -   Focused, measurable, time-bound, falsifiable … but not grounded.
    -   Nothing in the evidence suggests TikTok ads or 10,000 buyers. It’s just a guess.
-   Below the threshold (high uncertainty, strong grounding): “If we **sell samples at the campus gym** for two weeks, then at least 50 unique customers will buy at \$2.50.”
    -   Still focused, measurable, time-bound, falsifiable.
    -   Grounded because they’ve observed gym-goers choosing healthier options and showing interest in alternatives to soda.
:::

````{=html}
<!--
```{r}
#| label: fig-groundedness-threshold-example
#| fig-cap: "One example of high uncertainty and high grounding (selling at the campus gym) compared to an example of high uncertainty and low grounding (running TikTok ads)."
#| fig-width: 7
#| fig-height: 5

library(ggplot2)

# Make a small grid for plotting
df <- expand.grid(
  Groundedness = seq(0, 1, length.out = 100),
  Uncertainty = seq(0, 1, length.out = 100)
)

# Define threshold line: simple diagonal (Uncertainty = 1 - Groundedness)
df$Zone <- ifelse(df$Groundedness < df$Uncertainty, "Explore", "Hypothesize")

base_plot <- ggplot(df, aes(x = Groundedness, y = Uncertainty, fill = Zone)) +
  geom_tile(alpha = 0.7) +
  scale_fill_manual(
    values = c("Explore" = "#E6F7FA", "Hypothesize" = "#38B2AC")
  ) +
  geom_abline(slope = 1, intercept = 0, color = "#007BA7", linewidth = 2, linetype = "solid") +
  labs(
    x = "Grounding (Low → High)",
    y = "Uncertainty (Low → High)",
    fill = "Action" #,
#    title = "Applying the Groundedness Threshold for Hypotheses"
  ) +
  annotate("text", x = .25, y = .85, label = "Explore",
               color = "black", size = 5, fontface = "bold") +
  annotate("text", x = .25, y = .7, label = "H: TikTok ad: 10,000 buyers\n(above threshold — guessing zone)",
               color = "steelblue4", size = 4) +

  annotate("text", x = .65, y = .35, label = "Hypothesize",
               color = "black", size = 5, fontface = "bold") +
  annotate("text", x = .65, y = .2, label = "Campus gym test: 50 buyers\n(below threshold — grounded zone)",
               color = "black", size = 4) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", color = "#1A202C"),
    axis.title = element_text(face = "bold"),
    axis.text = element_blank(),   # alternative way (also removes values)
    legend.position = "top"
  )
base_plot

```
-->
